That’s a Big-Ass Chicken: A Mixed Methods Examination of Urban Backyard Chicken Husbandry

### 2025-03-08 BAC_Manuscript_Final.qmd
updated Qmd with fixes

---
title: "That’s a Big-Ass Chicken: A Mixed Methods Examination of Urban Backyard Chicken Husbandry"
shorttitle: "That's a Big-Ass Chicken"
author:
  - name: Karen Veronica Becerra
    corresponding: true
    orcid: 0009-0006-4967-0955
    email: kvbecerra@Uchicago.edu 
    affiliations:
      - name: The University of Chicago 
        department: Department of Psychology 
        address: 5848 S. University Avenue
        city: Chicago
        region: IL
        country: USA
        postal-code: 60637
        id: uchicago
    role:
      - conceptualization
      - formal anlaysis
      - methodology
      - project administration
      - software
      - visualization
      - editing 
  - name: Kristen M. Johnson
    orcid: 0009-0008-3757-5795
    email: kmjohnson25@uchicago.edu 
    affiliations:
      - ref: uchicago
    role:
      - conceptualization
      - data curation
      - formal analysis
      - methodology
      - project administration
      - software
      - writing
      - editing
  - name: Michael Moroz
    orcid: 0009-0000-9772-0001
    email: mrmoroz@uchicago.edu
    affiliations:
      - ref: uchicago
    role:
      - formal anlaysis
      - software
      - visualization
      - editing
abstract: "This study investigates the growing phenomenon of backyard chicken keeping in urban and suburban environments, a practice that has gained significant momentum amid rising commercial egg prices. Employing a mixed methods approach, we analyzed quantitative data from ______. Concurrently, we conducted thematic analysis of in-depth interviews with backyard chicken owners to explore perceived benefits and challenges. Our findings indicate that while initial setup costs average $487, households recover this investment within 14 months through egg production, with additional benefits including reduced food waste, improved garden soil quality, and reported increases in household well-being. Challenges identified include time commitment (averaging 28 minutes daily), navigating municipal regulations, and managing seasonal variations in productivity. Factor analysis revealed that success factors cluster around three dimensions: appropriate infrastructure, consistent maintenance routines, and community support networks. Ultimately, our results support the hypothesis that the benefits of backyard chicken keeping outweigh the challenges for most practitioners, though successful implementation depends on household-specific variables and community context. This research contributes to growing literature on urban agriculture while providing practical insights for potential chicken keepers, municipal policymakers, and sustainability advocates."
keywords: [chickens, backyard, eggs, pros and cons, bigger is better]
author-note:
  disclosures:
    conflict of interest: The authors have no conflicts of interest to declare.
  financial-support: "This study was supported by a grant from The International Society for Fowl Language."
  gratitude: "The authors are grateful to Eddy the Chicken for her endless inspiration and reluctant support."
impact-statement: "This study illustrates not only how, but also why one should raise big-ass backyard chickens. The results may be used by numerous individuals to convince their domestic partners (such as the second author's husband) to build a backyard coop and give in to the siren squawk of that sweet sweet chicken-raisin' lifestyle."
bibliography: Chickens.bib
format: 
  #apaquarto-docx: default 
  apaquarto-html: default    # remember to reverse commenting out other formats when turning project in
  apaquarto-pdf: default
floatsintext: true
documentmode: man
donotrepeattitle: true
mainfont: "Times New Roman"
fontsize: 12pt

@@

# Load necessary libraries
library(readr)
library(tidyverse)
library(ggplot2)
library(tidyverse)
library(flextable)
library(knitr)
library(kableExtra)
library(dplyr)
```

> *If you've ever seen a picture of Pechugas the Rooster (Fig. 1), your first words would invariably be "That's a Big-Ass Chicken!" And you would be correct.*

@@

Previous studies have largely focused either on large-scale commercial poultry operations or on historical agricultural practices [@bistSustainablePoultryFarming2024; @grzinicIntensivePoultryFarming2023; @jeniOverviewHealthChallenges2021].  While proponents extol the virtues of fresh eggs and animated garden assistants, detractors cite concerns ranging from zoonotic disease transmission to neighborhood noise pollution---a concern that anyone who has experienced a rooster's enthusiasm for announcing dawn can readily appreciate. Surprisingly, of those studies that do consider the costs and benefits of backyard chicken keeping, none that to the authors knowledge look into the phenomenon of the **Big-Ass Chicken** (BAC)[@ayalaReviewPathogenTransmission2020; @fischerDefenceBackyardChickens2019; @schindlerBackyardChickensFrontYard; @singhBackyardPoultryFarming2022]. This study aims to fill that obvious (and sizeable) gap.

@@

# Method
## Participants
### Chickens
We looked at 50 Gallus gallus domesticus chicks, starting from hatching.

### Humans
Semi-structured interviews were conducted with 50 urban backyard chicken keepers (32 female, 17 male, 1 non-binary) across three metropolitan areas (Eastern Seaboard, n=20; Midwest, n=15; Pacific Northwest, n=15). Participants had maintained backyard chickens for periods ranging from 6 months to 12 years (mean = 3.7 years). Flock sizes ranged from 2-15 birds (median = 4).

## Procedure
### Quantitative
We measured the body weights of the chickensat birth and every second day thereafter until day 20. They were also measured on day 21. There were four groups of chicks on different protein diets to test the effect of protein in 4 types of feed on early development of chicks. The 4 feeds are a control (normal diet) and 3 test feeds (10%, 20%, and 40% protein replacement).

## Measures
### Quantitative

### Qualitative
thematic analysis

```{r}
#| label: chicken_weight_growth_plot
#| echo: FALSE 
#| include: FALSE

```{r}
#| label: chicken_weight_growth-rate
#| echo: FALSE 
#| include: FALSE

## Ok now we have visualized the different growth trends of the various chickens. Now we probably want to look at how diet might factor into this trend. First let's calculate the mean growth rates of all the various chickens according to diet (rated 1 to 4)

print(dGrowRate)
```

```{r}
#| label: chicken_weight_growth-rate_plot
#| echo: FALSE 
#| include: FALSE

```{r}
#| label: chicken_weight_regression
#| echo: FALSE 
#| include: FALSE

## Eggs 
```{r}
#| label: chicken_egg_data
#| echo: FALSE 
#| include: FALSE

@@

# Save the chicken data to a CSV file
write.csv(chicken_data, "chicken_egg_weight_data.csv", row.names = FALSE)
```

```{r}
#| label: chicken_egg_size_weight
#| echo: FALSE 
#| include: FALSE

@@

```{r}
#| label: chicken-egg_summary_stats
#| echo: FALSE
#| include: FALSE

@@

```{r}
#| label: chicken-egg_correlation
#| echo: FALSE
#| include: FALSE

@@

```{r}
#| label: chicken-egg_correlation_plot
#| echo: FALSE

# Create a new column for average egg weight based on breed and egg size
chicken_data$Egg_Weight <- round(runif(nrow(chicken_data), 50, 80), 2)  # Random egg weights between 50 and 80 grams

@@

# Create a bar plot with APA styling
ggplot(egg_summary, aes(x = Breed, y = Average_Egg_Weight, fill = Egg_Size)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +  # Adjust bar width for APA style
  scale_fill_manual(values = c("Small" = "red", "Medium" = "yellow", "Large" = "green", "X-Large" = "blue")) +  # Custom colors
  labs(title = "Average Egg Weight by Chicken Breed and Egg Size", 
       x = "Chicken Breed", 
       y = "Average Egg Weight (grams)", 
       fill = "Egg Size") +
  theme_minimal(base_size = 12) +  # Set base font size
  theme(
    axis.title = element_text(size = 12),  # Axis titles in font size 12
    axis.text = element_text(size = 10),  # Axis text in font size 10
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Center and bold title
    legend.title = element_text(size = 12),  # Legend title font size
    legend.text = element_text(size = 10),  # Legend text font size
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    panel.grid.major = element_blank(),  # Remove major gridlines 
    panel.grid.minor = element_blank(),  # Remove minor gridlines
    panel.border = element_rect(color = "black", fill = NA)  # Add border
  )
```

@@

### 2. Self-Sufficiency and Food Security

#### 2.1 Control Over Food Supply
A dominant theme emerged around increased feelings of food security and self-reliance:

@@

### 3. Social and Community Dimensions

#### 3.1 Neighborhood Connections
Unexpectedly, chicken keeping frequently facilitated community building:

@@

### 4. Practical Challenges and Adaptations

#### 4.1 Time Commitment Reality
Time requirements emerged as the most frequently cited challenge (n=41, 82%):

@@

### 5. Psychological and Emotional Benefits

#### 5.1 Stress Reduction
An unexpected theme emerged around the stress-reducing properties of chicken keeping:

>>>>

---
title: "That’s a Big-Ass Chicken: A Mixed Methods Examination of Urban Backyard Chicken Husbandry"
shorttitle: "That's a Big-Ass Chicken"
author:
  - name: Karen Veronica Becerra
    corresponding: true
    orcid: 0009-0006-4967-0955
    email: kvbecerra@Uchicago.edu 
    affiliations:
      - name: The University of Chicago 
        department: Department of Psychology 
        address: 5848 S. University Avenue
        city: Chicago
        region: IL
        country: USA
        postal-code: 60637
        id: uchicago
    role:
      - conceptualization
      - formal anlaysis
      - methodology
      - project administration
      - software
      - visualization
      - editing 
  - name: Kristen M. Johnson
    orcid: 0009-0008-3757-5795
    email: kmjohnson25@uchicago.edu 
    affiliations:
      - ref: uchicago
    role:
      - conceptualization
      - data curation
      - formal analysis
      - methodology
      - project administration
      - software
      - writing
      - editing
  - name: Michael Moroz
    orcid: 0009-0000-9772-0001
    email: mrmoroz@uchicago.edu
    affiliations:
      - ref: uchicago
    role:
      - formal anlaysis
      - software
      - visualization
      - editing
abstract: "This study investigates the growing phenomenon of backyard chicken keeping in urban and suburban environments, a practice that has gained significant momentum amid rising commercial egg prices. Employing a mixed methods approach, we analyzed quantitative data from ______. Concurrently, we conducted thematic analysis of in-depth interviews with backyard chicken owners to explore perceived benefits and challenges. Our findings indicate that while initial setup costs average $487, households recover this investment within 14 months through egg production, with additional benefits including reduced food waste, improved garden soil quality, and reported increases in household well-being. Challenges identified include time commitment (averaging 28 minutes daily), navigating municipal regulations, and managing seasonal variations in productivity. Factor analysis revealed that success factors cluster around three dimensions: appropriate infrastructure, consistent maintenance routines, and community support networks. Ultimately, our results support the hypothesis that the benefits of backyard chicken keeping outweigh the challenges for most practitioners, though successful implementation depends on household-specific variables and community context. This research contributes to growing literature on urban agriculture while providing practical insights for potential chicken keepers, municipal policymakers, and sustainability advocates."
keywords: [chickens, backyard, eggs, pros and cons, bigger is better]
author-note:
  disclosures:
    conflict of interest: The authors have no conflicts of interest to declare.
    financial-support: "This study was supported by a grant from The International Society for Fowl Language."
    gratitude: "The authors are grateful to Edgar the Chicken for her endless inspiration and reluctant support."
impact-statement: "This study illustrates not only how, but also why one should raise big-ass backyard chickens. The results may be used by numerous individuals to convince their domestic partners (such as the second author's husband) to build a backyard coop and give in to the siren squawk of that sweet sweet chicken-raisin' lifestyle."
bibliography: Chickens.bib
format: 
  #apaquarto-docx: default 
  apaquarto-html: default    # remember to reverse commenting out other formats when turning project in
  apaquarto-pdf: default
floatsintext: true
documentmode: man
donotrepeattitle: true
mainfont: "Times New Roman"
fontsize: 12pt

@@

# Load necessary libraries
library(readr)
library(tidyverse)
library(ggplot2)
library(tidyverse)
library(flextable)
library(knitr)
library(kableExtra)
library(dplyr)
```
> *If you've ever seen a picture of Pechugas the Rooster (Fig. 1), your first words would invariably be "That's a Big-Ass Chicken!" And you would be correct.*

@@

Previous studies have largely focused either on large-scale commercial poultry operations or on historical agricultural practices [@bistSustainablePoultryFarming2024; @grzinicIntensivePoultryFarming2023; @jeniOverviewHealthChallenges2021].  While proponents extol the virtues of fresh eggs and animated garden assistants, detractors cite concerns ranging from zoonotic disease transmission to neighborhood noise pollution---a concern that anyone who has experienced a rooster's enthusiasm for announcing dawn can readily appreciate. Surprisingly, of those studies that do consider the costs and benefits of backyard chicken keeping, none that to the authors knowledge look into the phenomenon of the **Big-Ass Chicken (BAC)** [@ayalaReviewPathogenTransmission2020; @fischerDefenceBackyardChickens2019; @schindlerBackyardChickensFrontYard; @singhBackyardPoultryFarming2022]. This study aims to fill that obvious (and sizeable) gap.

@@

# Method
## Participants
### Chickens
```{r}
#| label: read_Gallus-Gallus_df
#| echo: FALSE 
#| include: FALSE

## Reading the data 
Gallus_Gallus = read.csv("Data_BigAssChicken/GallusGallusDomesticus.csv")

```
We looked at a total of **`r n_distinct(Gallus_Gallus$GallusID)`** Gallus gallus domesticus chicks, starting from hatching.

### Eggs
```{r}
#| label: chicken_egg_data
#| echo: FALSE 
#| include: FALSE

@@

# Save the chicken data to a CSV file
write.csv(chicken_data, "chicken_egg_weight_data.csv", row.names = FALSE)
```  
We looked at a total of **`r n_distinct(chicken_data$Eggs_Laid)`** eggs laid from **`r n_distinct(chicken_data$Chicken_ID)`** chickens.

### Humans
```{r}
#| label: urban-chicken-keepers_df
#| echo: FALSE 
#| include: FALSE
# Creating human data frame:
# Set seed for reproducibility
set.seed(123)

# Define the sample size
n <- 50

# Generate gender data (32 female, 17 male, 1 non-binary)
gender <- c(rep("Female", 32), rep("Male", 17), rep("Non-binary", 1))

# Generate location data (20 Eastern Seaboard, 15 Midwest, 15 Pacific Northwest)
location <- c(rep("Eastern Seaboard", 20), rep("Midwest", 15), rep("Pacific Northwest", 15))

# Generate experience data (randomly between 0.5 and 12 years)
experience <- round(runif(n, 0.5, 12), 1)

# Generate flock size data (randomly between 2 and 15 chickens)
flock_size <- sample(2:15, n, replace = TRUE)

# Create the data frame
urban_chicken_keepers <- data.frame(
  ID = 1:n,
  Gender = gender,
  Location = location,
  Experience = experience,
  Flock_Size = flock_size
)

# View the first few rows of the data frame
head(urban_chicken_keepers)
```
Semi-structured interviews were conducted with **`r n_distinct(urban_chicken_keepers$ID)`** urban backyard chicken keepers (`r sum(urban_chicken_keepers$Gender == "Female")` female, `r sum(urban_chicken_keepers$Gender == "Male")` male,  `r sum(urban_chicken_keepers$Gender == "Non-binary")` non-binary) across three metropolitan areas (Eastern Seaboard, n=`r sum(urban_chicken_keepers$Location == "Eastern Seaboard")`; Midwest, n=`r sum(urban_chicken_keepers$Location == "Midwest")`; Pacific Northwest, n=`r sum(urban_chicken_keepers$Location == "Pacific Northwest")`). Participants had maintained backyard chickens for periods ranging from 6 months to 12 years (mean = `r; Pacific Northwest, n=15). Participants had maintained backyard chickens for periods ranging from 6 months to 12 years (mean = `r mean(urban_chicken_keepers$Experience)`). Flock sizes ranged from 2-15 birds (median = `r median(urban_chicken_keepers$Flock_Size)`).

## Procedure
### Quantitative
We measured the body weights of the chickensat birth and every second day thereafter until day 20. They were also measured on day 21. There were four groups of chicks on different protein diets to test the effect of protein in 4 types of feed on early development of chicks. The 4 feeds are a control (average diet of <10% protein replacement is Diet 1) and 3 test feeds (10% is Diet 2, 20% is Diet 3, and 40% protein replacement is Diet 4).

## Measures
### Quantitative
We collected weight measurements using scales that are tiny and adorable for baby chickens/eggs and then also some considerably bigger scales for the BAC's.

### Qualitative
We applied thematic analysis to find relevant themes and then came to consensus on the affect valence of each response through the IOD (It's Obvious, Dummy) scale.

```{r}
#| label: chicken_weight_growth_plot
#| echo: FALSE

```
Figure 1 shows the different growth trends of the various chickens.
```{r}
#| label: chicken_weight_growth-rate
#| echo: FALSE 
#| include: false
## Ok now we have visualized the different growth trends of the various chickens. Now we probably want to look at how diet might factor into this trend. First let's calculate the mean growth rates of all the various chickens according to diet (rated 1 to 4)

print(dGrowRate)
```
Looking at diet's effect on growth rates, we found that protein content of 20% had the most positive effect on chicken growth.
```{r}
#| label: chicken_weight_growth-rate_plot
#| echo: FALSE

```
As shown in Figure 2 we see that diet 3 has chickens with the highest mean growth rates (per day). Diet 4 comes in second, then diet 2, then diet 1.
```{r}
#| label: chicken_weight_regression
#| echo: FALSE 
#| include: false

```
We found that both time and diet are significant predictors of weight. Meaning a one unit increase in time (which actually means going from 0 to 2, or 2 to 4, etc) corresponds with an 8.77 gram increase in weight. In addition, a one unit increase in diet (meaning going from diet 1 to 2, 2 to 3, or 3 to 4) corresponds with an 11.78 gram increase in weight. So the interpretation is that if using diet 4 as opposed to diet 1, you would see on average approximately a 33 gram weight increase.

## Eggs 
As shown in Table 1, we found that a component of our hypothesis was true: Big-Ass Chickens lay a Big-Ass-load of eggs.
```{r}
#| label: chicken_egg_size_weight
#| echo: FALSE 
#| include: false

@@

```{r}
#| label: chicken-egg_summary_stats
#| echo: FALSE

@@

As shown in Table 2 and contrary to our hypothesis, however, there is a negative correlation between chicken weight and egg production within the Big-Ass Chicken demographic. A Pearson correlation was conducted to examine the relationship between chicken weight and eggs laid. The results indicated a strong, negative correlation between the two variables, r(30) = -0.93, p = .02. This suggests that as chicken weight increases, the number of eggs laid tends to decrease.
```{r}
#| label: chicken-egg_correlation
#| echo: FALSE

@@

However, as you can see in Figure 4, larger BAC's produced larger eggs, so really our main hypothesis of more chicken = more egg is like actually totally right when you think about it. It's not just about the number of eggs, but also the quantity of goo within each egg. Put another way, it's not just the size that matters, but how you use it.... it being the egg goo.
```{r}
#| label: chicken-egg_correlation_plot
#| echo: FALSE
#| fig-cap: Average Egg Weight by Chicken Breed and Egg Size
# Create a new column for average egg weight based on breed and egg size
chicken_data$Egg_Weight <- round(runif(nrow(chicken_data), 50, 80), 2)  # Random egg weights between 50 and 80 grams

@@

# Create a bar plot with APA styling
ggplot(egg_summary, aes(x = Breed, y = Average_Egg_Weight, fill = Egg_Size)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +  # Adjust bar width for APA style
  scale_fill_manual(values = c("Small" = "red", "Medium" = "yellow", "Large" = "green", "X-Large" = "blue")) +  # Custom colors
  labs(x = "Chicken Breed", 
       y = "Average Egg Weight (grams)", 
       fill = "Egg Size") +
  theme_minimal(base_size = 12) +  # Set base font size
  theme(
    axis.title = element_text(size = 12),  # Axis titles in font size 12
    axis.text = element_text(size = 10),  # Axis text in font size 10
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Center and bold title
    legend.title = element_text(size = 12),  # Legend title font size
    legend.text = element_text(size = 10),  # Legend text font size
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    panel.grid.major = element_blank(),  # Remove major gridlines 
    panel.grid.minor = element_blank(),  # Remove minor gridlines
    panel.border = element_rect(color = "black", fill = NA)  # Add border
  )
```

@@

### 2. Self-Sufficiency and Food Security
#### 2.1 Control Over Food Supply
A dominant theme emerged around increased feelings of food security and self-reliance:

@@

### 3. Social and Community Dimensions
#### 3.1 Neighborhood Connections
Unexpectedly, chicken keeping frequently facilitated community building:

@@

### 4. Practical Challenges and Adaptations
#### 4.1 Time Commitment Reality
Time requirements emerged as the most frequently cited challenge (n=41, 82%):

@@

### 5. Psychological and Emotional Benefits
#### 5.1 Stress Reduction
An unexpected theme emerged around the stress-reducing properties of chicken keeping:
